Estimating the Initial Susceptible Fraction in Dengue Dynamics

### 2014-09-29 draft_lm1.tex
Improvements of Sep 29th 2014

confers permanente immunity within serotypes but with only short-lived 
cross-immunity. Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity, Thirdly, under-reporting is 
a serious issue in Brazil. Duarte and Franca\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting in Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the cases go unreported. Lastly, demography 
and migrations affect the number of susceptible in ways which are not easy to
determine.

@@

\section*{Methods}
\subsection*{The data}

The data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil.
From this series, we calculated the effective reproduction number, $R_t$, according to the expression given by Nishiura et al.~\cite{nishiura}.
The method assumes the disease counts $Y(t)$ are Poisson random variables such that $R_t$ is 
\[ R_t = \left(\frac{Y(t + 1)}{Y(t)}\right)^{1/n}\]
where $n$ is the ratio between the length of reporting interval and the mean generation time of the disease.
In this paper we obtain the estimates of $R_t$ using a moving window of three weeks, which roughly corresponds to the generation time of dengue, and thus $n = 1$.
We provide more detail on the estimation and confidence/credibility intervals in Text S1.

@@

A Susceptible-Infectious-Removed (SIR) model is proposed to model dengue dynamics.
In the traditional formulation of the model, transmission is governed by a transmission rate $\beta$ and recovery happens at a rate $\tau$.
As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation}
 \label{eq:effbeta}
 \beta(t) = R_t	\cdot\tau
\end{equation}

@@

\subsection*{Bayesian parameter estimation}
For the fitting procedure the time series were normalized to lie on the $[0,1]$ interval, therefore avoiding scaling issues.
% Results and Discussion can be combined.xx
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the estimation of $R(t)$.

>>>>

confers permanente immunity within serotypes but with only short-lived 
cross-immunity. Secondly, there are a larger number of asymptomatic cases in 
any epidemic, which nevertheless acquire immunity, Thirdly, under-reporting is 
a serious issue in Brazil. Duarte and Fran\c{c}a\cite{duarte_data_2006}, estimated 
the sensitivity of dengue reporting in Belo-Horizonte, Brazil to be of 63\%, 
meaning that aproximately 37\% of the cases go unreported. Lastly, demography 
and migrations affect the number of susceptible in ways which are not easy to
determine.

@@

\section*{Methods}

\subsection*{The data}

The data used to fit the model consists of weekly incidence from 1996 to 2014 
in the city of Rio de Janeiro, Brazil.
From this series, we calculated the effective reproduction number, $R_t$, 
according to the expression given by Nishiura et al.~\cite{nishiura}.
The method assumes the disease counts $Y(t)$ are Poisson random variables such 
that $R_t$ is 
\[ R_t = \left(\frac{Y(t + 1)}{Y(t)}\right)^{1/n}\]
where $n$ is the ratio between the length of reporting interval and the mean 
generation time of the disease.
Here we obtain the estimates of $R_t$ using a moving window of three weeks, 
which roughly corresponds to the generation time of dengue, and thus $n = 1$.

We provide more detail on drawing posterior inference for $R_t$ in Text S1.

@@

A Susceptible-Infectious-Removed (SIR) model is proposed to model dengue dynamics.
In the traditional formulation of the model, transmission is governed by a 
transmission rate $\beta$ and recovery happens at a rate $\tau$.

As the epidemic progresses, the effective transmission  rate changes and is
\begin{equation} 
 \label{eq:effbeta}
 \beta(t) = \frac{R_t\cdot\tau}{S_0}
\end{equation}

@@

\subsection*{Bayesian parameter estimation}

For the fitting procedure the time series were divided into epidemic windows 
that corresponded to significant raises in incidence and normalized to lie on 
the $[0,1]$ interval.

Also, as the numerical integration routine needs $\beta(t)$ to be continuous, we 
used linear interpolation to obtain values of $R_t$ for any time point.

We take a Bayesian approach to the estimation of $S_0$ for each window.
For a given interval $j = \{ t_j^{\text{start}}, t_j^{\text{end}} \} $ we 
observe an incidence time series $\mathbf{Y_{j}}$.
We are thus interested in the posterior distribution
\begin{equation}
 \label{eq:S0post}
 p(S_{0j}|\mathbf{Y_{j}}) \propto l(\mathbf{Y_{j}}|S_{0j}, R_t, m,  \tau 
)\pi(S_{0j}) 
\end{equation}
The likelihood $l(\mathbf{Y_{j}}|\cdot)$ is assumed to be a Normal distribution 
with fixed variance $\sigma^2$.
In this estimation procedure we kept $R_t$ fixed at fixed at the posterior mean 
obtained as described in TextS1 and fixed $\tau = 1/7\: \text{days}^{-1}$.
% \subsubsection*{Priors}
To complete the inference, we need to specify prior distributions for the 
parameters of interest.
We place a flat $\text{Beta}(1, 1)$ prior on $S_{0j}\:\forall j$
% 
% \begin{align}
%  S_{0j} &\sim Beta(1, 1) \\
%        m &\sim \\
% \end{align}

% \subsubsection*{Posterior approximation}

To approximate the posterior in~(\ref{eq:S0post}) we use Markov chain Monte 
Carlo techiniques implemented in the Bayesian inference with Python 
(BIP)~\cite{pone2011} available at 
\url{http://code.google.com/p/bayesian-inference/}.
BIP uses a Differential Evolution Adaptive Metropolis (DREAM)~\cite{vrugt2008} 
scheme that efficiently samples from high-dimensional joint distributions using 
multiple adaptive chains running in parallel with delayed rejection.
The chains were run until $5000$ samples were obtained after discarding $500$ 
burn-in samples.
Convergence of the parallel chains was verified at every 100 iterations by the 
calculation of the Gelman-Rubins' R convergence diagnostic~\cite{brooks1998}.
% Results and Discussion can be combined.
\section*{Results}
% We only support three levels of headings, please do not create a heading level below \subsubsection.
%\subsection*{Subsection 1}
\section*{Discussion}
% Do NOT remove this, even if you are not including acknowledgments.
\section*{Acknowledgments}
LMC is grateful to Dr. Leonardo Bastos for useful discussions on the 
posterior inference for $R(t)$.
